\textbf{NBA Party}

### 2017-07-06 cardinal_analytx_presentation/nightlife_and_sports.tex
The first couple slides are *starting* to look more polished.

\begin{frame}   \frametitle{What is the problem?}
  \begin{block}{Do pro athletes party to the detriment of their performance?}
    Anecdotally, they do party.\footnote{Duffy ('17), Cestone ('17), Duffy ('16), 
      Price ('15), Concepcion ('15), Miller ('15), Campbell ('15), Ball ('12).}

    \vspace{12pt}It's not unreasonable to think that what happens off the court affects game-outcomes, but!
    \begin{itemize}       
      \item Can we even measure a hangover effect?
      \item Are there systematic tendencies?
      \end{itemize}
  \end{block} 
\end{frame}

@@

\begin{frame}   
  \frametitle{Whether players actually party is a latent variable}
  \begin{block}{We can't observe whether players party.}
    Instead, we will consider a quasi-experiment where players are exposed to nightlife. 
    Motivating questions:
    \vspace{12pt} 
    \begin{itemize}       \item Do teams play worse the day after visiting a ``party city''?
      \item How do we even determine whether a city has nightlife?    
      \item What does it mean to ``play worse''? \end{itemize}
  \end{block}

  \vspace{12pt}We define ``play worse'' relative to bookmaker/public expectations.
\end{frame}

@@

    Apparently, a lot of people.

    \vspace{12pt}Even aside from bettors, knowing the causal relation between partying and next-day performance might inform
    \begin{itemize}       
      \item how coaches set team-rules, 
      \item how travel schedules are determined, as well as
      \item which players to bench, and when.
      \end{itemize}
  \end{block}
\end{frame}

\begin{frame}   
  \frametitle{Is the problem hard?}

  \begin{block}{Finding causal-mechanisms which significantly determine game-outcomes is difficult.}       It's not obvious whether we will find an effect.

      \vspace{12pt}After all, we can't reliably observe if players party, let alone when.

      Even if we could, can we exploit this in the marketplace?   \end{block}

\begin{frame}
  \frametitle{Previous research - NBA player shooting average}     
  NBA Players under-perform proportion of shots made on Sundays.\footnote{Chase, ('15).}
    \begin{itemize}       
    \item 1,092 of 2,035 NBA players who took $\geq 1$ shot on Sundays under-perform 
      relative to their own season-long average; $p=0.001$.
      \item Effect size is not considered.
      \item Density of support is questionable.\footnote{What if we miss the only shot we took on Sunday, is that relevant?}     \end{itemize}
\end{frame}

\begin{frame}   \frametitle{Prev. research - likelihood of meeting the spread in NBA}
  Teams under-perform bookmakers expectations on Sunday afternoons when on the road.\footnote{Ezekowitz, ('14).}
\vspace{12pt}
    \begin{itemize}       
    \item Only 129 out of 284 games did the away team meet the spread, 
      using 2006-14 data; $p=0.1089$.
    \item Not quite significant at the 90\% confidence level.     
    \item If we bet \$100 on each Sunday game, we'd only profit \$1,310.\footnote{$(284 - 129) \times 100 + 129 \times (-110) = 1,310$.} 
    \end{itemize}
  \end{frame}

@@

    A team is expected to meet the spread $\frac{1}{2}$ the time.\footnote{Dubner and Levitt, ('06). Losers pay 10\% fee on top of wagers; a balanced book guarantees 5\% gain for the house-bookmaker.}
  \end{block}
  \begin{block}{Bookmaker money-lines (MLB)}     The odds that a team will beat their opponent.

@@

\begin{frame}   \frametitle{Identification}
    If next day opponent is uncorrelated with visitation to a party city,
    then looking at the \emph{actual outcome} of the game $\leadsto$ identification.

@@

    Weaker, still: if odds of beating the spread uncorrelated with previous game location
    $\leadsto$ identification.\footnote{Under an efficient market, this assumption should hold.}     \end{frame}

\begin{frame}   \frametitle{Our solution: a quasi-experiment}
  Consider games played back-to-back within 24 hours.

  \vspace{12pt}
  Examine whether spread is met as a function of nightlife index for lagged game location.

  \vspace{12pt} (Alcohol and loss of sleep are well correlated with nightlife; both have been shown to have negative effects on sports performance.)\footnote{Shirreffs and Maughan ('06), Barnes ('14), Pilcher and Huffcutt ('96).} \end{frame}

\begin{frame}   \frametitle{Sports: NBA and MLB}
  \begin{block}{We perform the analysis for both NBA and MLB}     
    Observe a team traveling to a city with active nightlife.
    \vspace{12pt}
    \begin{itemize}       \item For NBA teams playing another game in the next 24 hours, are they less likely to meet the points-spread?
      \item For MLB teams, after conditioning on the odds of winning, are they less likely to beat their opponent?     \end{itemize}
    \vspace{12pt}\begin{block}{If we yield a causal model, we may realize a betting schema.}     \end{block}
  \end{block} 
\end{frame}

@@

\begin{frame}   \frametitle{Further motivation for conditioning on the spread}
  \begin{block}{With some correlation between next-day opponent, this further motivates conditioning
  on bookmakers' expectations.}
  \begin{itemize}     \item If bookkeepers aware of hangover effects, our variable of interest may not appear significant, in spite of there being a true effect.

>>>>

\begin{frame}   \frametitle{What is the problem?}
  \begin{block}{Do pro athletes party to the detriment of their performance?}     It's not unreasonable to think that what happens off the court affects game-outcomes, but!
    \begin{itemize}       
      \item Can we even measure these phenomena?
      \item Are there systematic tendencies?
    \end{itemize}       \end{block}
\end{frame}

@@

\begin{frame}   
  \frametitle{Why is this problem hard?}
  \begin{block}{Partying is a latent variable.}
    Instead, we proxy for the possibility of athletes partying by looking at 
    play following games in party cities. Our question becomes...
  \end{block}
  \vspace{12pt}\begin{block}{Do teams exhibit a decline in performance the day following a game in a city with active nightlife?}     I.e. we are concerned with ``hangover effects''.
  \end{block}
  \vspace{12pt}\begin{block}{Part of the question is determining a measure for nightlife. }
    We colloquially call cities with high levels of nightlife ``party cities''.   \end{block}
\end{frame}

  %   \vspace{12pt} 
  %   \begin{itemize}       \item Do teams play worse the day after visiting a ``party city''?
  %     \item How do we even determine whether a city has nightlife?    
  %     \item What does it mean to ``play worse''? \end{itemize}

  % \vspace{12pt}We define ``play worse'' relative to bookmaker/public expectations.

@@

    Apparently, a lot of people. Aside from bettors...
  \end{block}

  \vspace{12pt}\begin{block}{The causal relationship might inform more optimal strategy.}     For example,
    % Even aside from bettors, knowing the causal relation between % partying and next-day performance might inform
    \begin{itemize}       
      \item how coaches set team-rules, 
      \item how managers determine travel schedules, as well as
      \item which players to bench, and when.\footnote{Are players more likely to get injured the next day as a result of partying?}
    \end{itemize}
  \end{block}
\end{frame}

% \begin{frame}   
%   \frametitle{Is the problem hard?}

%   \begin{block}{Finding causal-mechanisms which significantly determine % game-outcomes is difficult.}       It's not obvious whether we will find an % effect.

%       \vspace{12pt}After all, we can't reliably observe if players party, let % alone when.

%       Even if we could, can we exploit this in the marketplace?   \end{block}

% \end{frame}

@@

    A team is expected to meet the spread $\frac{1}{2}$ the time.\footnote{Dubner and Levitt, ('06). Losers pay 10\% fee on top of wagers; a balanced book guarantees 5\% gain for the house-bookmaker.}
  \end{block}
  \vspace{12pt}\begin{block}{Bookmaker money-lines (MLB)}     The odds that a team will beat their opponent.

@@

\begin{frame}   \frametitle{Our solution: a quasi-experiment.}
  \begin{block}{Systematic deviations represent unexplained mechanisms.}     We construct a model which attempts to estimate the causal effect of visiting a ``party city''
    on subsequent day performance, as measured by the odds of meeting the spread.       \end{block}

  \begin{block}{}     We only consider the hangover effect on games played back-to-back within 24 hours of each other.   \end{block} \end{frame}

\begin{frame}   \frametitle{Identifying variation}
  \begin{block}{Explain performance as a function of prev. day's nightlife index.}     If next day opponent is uncorrelated with visitation to a party city,
    then looking at the \emph{actual outcome} of the game $\leadsto$ identification.

@@

    Weaker, still: if odds of beating the spread uncorrelated with previous game location
    $\leadsto$ identification.\footnote{Under an efficient market, this assumption should hold.}       \end{block}
\end{frame}

\begin{frame}
  \frametitle{Previous research - NBA player shooting average}
  \begin{block}{Empirical evidence: players party to the detriment of the game.}
  NBA Players under-perform proportion of shots made on Sundays.\footnote{Chase, ('15).}
  \end{block}
    \begin{itemize}       
    \item 1,092 of 2,035 NBA players who took $\geq 1$ shot on Sundays under-perform 
      relative to their own season-long average; $p=0.001$.
      \item Effect size is not considered.
      \item Density of support is questionable.\footnote{What if we miss the only shot we took on Sunday, is that relevant?}     \end{itemize}
\end{frame}

\begin{frame}   \frametitle{Prev. research - likelihood of meeting the spread in NBA}
  \begin{block}{Empirical evidence: away teams under-perform on Sundays.}
  Teams under-perform bookmakers expectations on Sunday afternoons when on the road.\footnote{Ezekowitz, ('14).}
    \begin{itemize}       
    \item Only 129 out of 284 games did the away team meet the spread, 
      using 2006-14 data; $p=0.1089$.
    \item Not quite significant at the 90\% confidence level.     
    \item If we bet \$100 on each Sunday game, we'd only profit \$1,310.\footnote{$(284 - 129) \times 100 + 129 \times (-110) = 1,310$.} 
    \end{itemize}
  \end{block}
  \end{frame}

% \begin{frame}   \frametitle{Our solution: a quasi-experiment}
%   Consider games played back-to-back within 24 hours.

%   \vspace{12pt}
%   Examine whether spread is met as a function of nightlife index for lagged game % location.

%   \vspace{12pt} (Alcohol and loss of sleep are well correlated with nightlife; % both have been shown to have negative effects on sports % performance.)\footnote{Shirreffs and Maughan ('06), Barnes ('14), Pilcher and % Huffcutt ('96).} \end{frame}

\begin{frame}   \frametitle{Sports: NBA and MLB}
  \begin{block}{We perform the analysis for both NBA and MLB}     
    If we observe a team traveling to a city with active nightlife:
    \vspace{12pt}
    \begin{itemize}       \item For NBA teams playing another game in the next 24 hours, are they less likely to meet the points-spread?
      \item For MLB teams, after conditioning on the odds of winning, are they less likely to beat their opponent?     \end{itemize}
    \vspace{12pt}\begin{block}{If we yield a causal model, we may realize a betting schema.}     
\end{block}
  \end{block} 
\end{frame}

@@

\begin{frame}   %\frametitle{Further motivation for conditioning on the spread}
  \begin{block}{With some correlation between next-day opponent, this further motivates conditioning
  on bookmakers' expectations.}
  \begin{itemize}     \item If bookkeepers aware of hangover effects, our variable of interest may not appear significant, in spite of there being a true effect.
